---
title: "Untitled"
output: html_document
date: "2024-08-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=F}
library(dplyr)
library(glm2)
library(glmnet)
library(devtools)
library(pak)
library(mice)
library(stringr)
library(umap)
library(caret)
library(plotly)
```

                                                                                                                  

### Reading data into a dataframe 
```{r}
DNAm <- data.frame(read.csv("BGSMethylation.csv", header = T))
```


```{r}
(sum(is.na(DNAm))/(nrow(DNAm)*(ncol(DNAm))))*100
```
There is therefore 0.24% missing values in the dataset 

Efficiency resource https://stackoverflow.com/questions/53409246/object-size-increases-hugely-when-transposing-a-data-frame
running parallel computing for mice: https://stackoverflow.com/questions/24040280/parallel-computation-of-multiple-imputations-by-using-mice

```{r}
rownames(DNAm) <- DNAm[,1] #adding row names also adds the column names  
DNAm <- DNAm[,-1] #removing the rowname column from the dataset as it is now a duplicate
DNAm <- as.matrix(DNAm) #converting data frame to double type matrix due to space efficiency 8Gb -> 1Gb
DNAmTr <- t(DNAm) #transposing data set
rm(DNAm) #removing OG data to free up space
```
 

I then replaced the missing values in the data with the column mean value
```{r}
#for (i in 1:ncol(DNAmTr)){
#  DNAmTr[is.na(DNAmTr[,i]), i] <- mean(DNAmTr[DNAmTr[,2] == DNAmTr[i,2], i], na.rm = T)
#}
```

```{r}
(sum(is.na(DNAmTr))/(nrow(DNAmTr)*(ncol(DNAmTr))))*100
```






```{r}
#format(object.size(DNA), units = "MB")
#format(object.size(DNAmTr), units = "MB")
```

### Preparing the age, sex data into double type data with string rownames in a matrix 

```{r}
#Reading in data set and removing the 'twin' variable as all values are 0 therefore no twins in the data set
DNAmINFO <- data.frame(read.csv("BGSInfo.csv", header = T))  %>% select(-c(X, Twin))
rownames(DNAmINFO) <- DNAmINFO$ID 
DNAmINFO <- DNAmINFO[,-1]

#Changing the sex variable to binary 
DNAmINFO[DNAmINFO$Sex == "M"] <- DNAmINFO$Sex[DNAmINFO$Sex == "M"] <- 1
DNAmINFO[DNAmINFO$Sex == "F"] <- DNAmINFO$Sex[DNAmINFO$Sex == "F"] <- 0

DNAmINFO$Sex <- as.double(DNAmINFO$Sex)
DNAmINFO$Age <- as.double(DNAmINFO$Age)

sapply(DNAmINFO, typeof) #checking the variab;e type to ensure a double type matrix

DNAmINFO <- as.matrix(DNAmINFO)
```



### Merging the two dataframes to have age and DNAm data in the same dataframe


```{r}
DNA <- merge(DNAmINFO, DNAmTr, by = "row.names", all = T) #merging changes 
row.names(DNA) <- DNA[,1]
DNA <- DNA[,-1]
DNA <- as.matrix(DNA)
#(DNAmINFO)
rm(DNAmTr)
gc() #to ensure the space is freed up by R 
```



```{r}
for (i in DNA) {
    for (j in DNA[, i]) {
        DNA[j, i] <- mean(DNA[DNA[, 2] == DNA[j, 2], i],  na.rm = TRUE)
    }
   
}
```

```{r}
typeof(DNA)
```

```{r}
DNAfull <- data.frame(DNA) %>% 
        group_by(Sex) %>% 
        mutate(., function(x) replace(x, is.na(x), mean(x, na.rm = TRUE)))
```


```{r}
DNAfull <- data.frame(DNA) %>% group_by(Sex) %>% 
  mutate()
```


```{r}
DNAfull <- data.frame(DNA) %>% group_by(Sex) %>% mutate_all(~ replace_na(., mean(., na.rm = TRUE)))
```



```{r}
for (i in length(DNA)) {
  DNA[,i]<-ave(DNA[,i],DNA[,1],FUN=function(x) 
    ifelse(is.na(x), mean(x,na.rm=TRUE), x))
  
}
```


```{r}
colnames(DNA)[1:5]
```


```{r}
sum(is.na(DNA))
```


### Splitting the data into test and train sets

```{r}
set.seed(123)
Index <- sample(nrow(DNA), size = nrow(DNA) * 0.8, replace = F)

TrainDNA <- DNA[Index,]
TestDNA <- DNA[-Index,]
```



```{r}
#write.csv(TrainDNA, "/Users/katee/OneDrive/Desktop/Industrial project/TrainDNA.csv")
#write.csv(TestDNA, "/Users/katee/OneDrive/Desktop/Industrial project/TestDNA.csv")
#write.csv(DNA, "/Users/katee/OneDrive/Desktop/Industrial project/DNA.csv")
```





```{r}
y.TrainDNA <- TrainDNA[,2, drop = F] # drop = F keeps the row names to stop it being converted to a vector
x.TrainDNA <- TrainDNA[,-2]
```

```{r}
y.TestDNA <- TestDNA[,2, drop = F] # drop = F keeps the row names to stop it being converted to a vector
x.TestDNA <- TestDNA[,-2]
```


```{r}
#rm(DNA)
rm(TestDNA)
rm(TrainDNA)
gc()
```



### Training elastic net models

Below i am training 11 different models with different alpha values 
```{r}
models <- list()

for(i in 0:10) {
  modelName <- paste0("alpha", i/10)
  models[[modelName]] <- cv.glmnet(x.TrainDNA, y.TrainDNA, type.measure="mse", alpha = i/10, family="gaussian")
}     
```

### Testing the accuracy of the trained models 

```{r}
accuracy <- data.frame()
for (i in 0:10) {
  modelName <- paste0("alpha", i/10)

  predictions <- predict(models[[modelName]], 
                         s = (models[[modelName]]$lambda.1se), new = x.TestDNA)
  
  mse <- mean((y.TestDNA - predictions)^2)
  temp <- data.frame(alpha = i/10, mse= mse, modelName = modelName)
  accuracy <- rbind(accuracy, temp)
  
}
```

```{r}
accuracy
```



### UMAP dimensionality reduction

```{r}
custom.config = umap.defaults # Set of configurations
custom.config$min_dist = 0.15 # change min_dist and n_neighbors
custom.config$n_neighbors = 10
custom.config$n_components = 16
```




```{r}
DNAumap <- umap(DNA[,-2], config=custom.config)
```

```{r}
df <- data.frame(DNAumap$layout[,1], DNAumap$layout[,2], DNA[,1], DNA[,2])
colnames(df) <- c("X","Y", "Gender", "Age")
umapMat <- as.matrix(df[-3], header = T)
Gender <- factor(df$Gender, levels=c(0,1), labels=c("Female", "Male")) 
ggplot(umapMat, aes(x =X, y= Y, color = Age, shape = Gender ))+ geom_point()  + labs(title="DNA umap") + scale_shape_discrete(breaks = ~ .x[!is.na(.x)])
```


```{r}
custom.config = umap.defaults # Set of configurations
custom.config$n_components = 3 # change min_dist and n_neighbors

DNAumap3d = umap(DNA[,-2], config=custom.config)
```

```{r}

dfc <- data.frame(DNAumap3d$layout[,1], DNAumap3d$layout[,2], DNAumap3d$layout[,3], DNA[,1], DNA[,2])
colnames(dfc) <- c("X","Y", "Z", "Gender", "Age" )
dfc$Gender <- factor(dfc$Gender, levels=c(0,1), labels=c("Female", "Male")) 
plot_ly(dfc, x=dfc$X, y=dfc$Y, z=dfc$Z, color = ~Age, symbol = ~Gender, symbols = c('square', 'circle'), type="scatter3d", mode="markers")
```



https://jtr13.github.io/cc21fall1/efficient-dimension-reduction-with-umap.html





### PCA dimensionality reduction

```{r}
DNApca <- prcomp(DNA[,-2], scale=T)
summary(DNApca)$importance
```

```{r}
df$Gender <- factor(df$Gender, levels=c(0,1), labels=c("Female", "Male")) 

dfpca <- data.frame(DNApca$x[,1], DNApca$x[,2], factor(DNA[,1], levels=c(0,1), labels=c("Female", "Male")) , DNA[,2])
colnames(dfpca) <- c("PCA1", "PCA2", "Gender", "Age") 


ggplot(dfpca, aes(x = PCA1, y= PCA2, color = Age, shape = Gender)) + geom_point()  + labs(title="DNA pca")
```

```{r}
DNApca <- prcomp(DNA[,-2], scale=T, tol = 2)
summary(DNApca)$importance
```



### UMAP neural network 

```{r}

umapData <- as.matrix(merge(DNAmINFO, as.matrix(DNAumap$layout), by = "row.names", all = T))
gc() #to ensure the space is freed up by R 
```







### PCA neural network 


```{r}
pcaData <- as.matrix(merge(DNAmINFO, as.matrix(DNApca$importance), by = "row.names", all = T))

DNApca$loadings
```






















